Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2019-10-16 Moral_Communities_and_Crime_v1.tex
Initial submission to Applied Spatial Analysis and Policy

This concern is not iddle: the endemic malady of violent crime
represents an important factor that threatens to derail sustainable
development goals. Violent crime, furthermore, tends to be more accute
in those regions most in need of development. On the one hand, violent
crime represents an economic and social drag (Becker and Kassouf 2017).
The impacts are multifaceted, since they affect people's well-being,
either through the loss of human life, mental health, and limitations in
the right to public spaces (Doran and Burgess 2011), or through
disturbances in schooling and academic achievement (United Nations
Office on Drugs and Crime 2019b). In turn, these negative effects
combine to create the unfortunate conditions that tend to breed criminal
behavior, thus creating a vicious cycle of economic disadvantage and
violence (United Nations Office on Drugs and Crime 2019b). It is not
surprising, given these pecularities, that researchers have joined calls
for increased attention to the study of the patterns of violent crime in
Low and Middle Income Countries (LMIC; Murray, Cerqueira, and Kahn
2013).

Added to the scenario, formal social control institutions in countries
like Brazil leave something (or much) to be desired: while in practice
they should inhibit criminal behavior, they suffer from deep
deficiencies that end up reducing the deterrent power of the justice
system. Serious institutional problems include the inefficiency of the
police, the lack of national legislation, the glacial pace of judicial
processes, and the weak situation of the prison system in the country
(Menezes et al. 2013). This makes it even more urgent to understand the
role of \emph{informal social controls}, i.e., the ability of community
organizations such as schools, clubs, and neighborbood associations to
suppress crime by strengthening the capabilities of neighbors to control
inappropriate behavior (Groff 2015).

In order to understand the factors that can potentially help to deter
crime, beyond formal control, it is important to identify empirical
regularities. Criminological factors include concrete elements (such as
te presence of arms or drugs, or elements of the built environment), and
also figurative factors, which include the soci costs of deviant
behavior as well as family supervision. Accordingly, a number of studies
have investigated various aspects of the environment and neighborhood
design (e.g., Foster, Giles-Corti, and Knuiman 2010; He, Páez, and Liu
2017; Loukaitou-Sideris et al. 2001), whereas other studies have focused
on exposure to environmental attributes that signal weakened norms, such
as liquor and tobacco outlets (e.g., Brower and Carroll 2007; Deryol et
al. 2016; Lipton et al. 2008; Quick, Law, and Luan 2017).

@@

With the above considerations in mind, the objective of this paper is to
investigate whether and how places of workship correlate spatially with
criminal events, or in other words, to investigate whether their signals
as moral communities have a discernible spatial dimension. The case
study is the city of Recife in the state of Pernambuco, in Brazil's
Northeast. Recife is a large and important metropolitan area in a
historically poor region, and afflicted by high levels of violent crime,
having the dubious honor of being one of five state capitals with the
highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

@@

An early effort to theorize the effect of religion on criminal behavior
was the \emph{hellfire hypothesis} of Hirschi and Stark (1969). The
focus of this hypothesis is the threat of extratemporal (and possibly
eternal) punishment, and how this threat can deter believers from
commandment-breaking (also see Pascal's wager). Accordingly, Hirsch and
Stark (1969) posit that negative correlations between crime and religion
are a consequence of a sense of commitment with normative values, a
commitment that is ritually reinforced in a regular fashion (e.g., the
liturgical rite of peace in Catholic Mass). Rohrbaugh and Jessor (1975),
for instance, note that by attributing to divinity (or some other
supra-terrenal entity) the supreme force of punishment, religion helps
to build a sense of obduracy, or ``hardening'' against temptation.

@@

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plaussibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization theory are
thought to occur at relatively small geographical scales.

@@

Using quadrats of equal size is a convenient simplication, but in
principle the areas could be different - in which case the count of
events would be proportional to the area of the quadrat under
homogeneity. An interesting variation of this technique is conditional
quadrat counts, whereby the partition of regions
\(\mathcal{A}_1,\cdots,\mathcal{A}_m\) is done to reflect an underlying
covariate of interest, say \(Z\). By introducing a covariate as a
partitioning criterion, it becomes possible to calculate estimators of
the intensity of quadrats at different levels of the value of the
covariate. For exploratory purposes, we can plot average intensity by
quadrat, and compare to the global intensity.

@@

\caption{\label{tab:table-descriptive-statistics}\label{tab:descriptive-statistics}Characteristics of Violent Crime in Recife, July, 2008 - June, 2009}
\centering
\begin{tabular}[t]{lr}
\toprule
 & Percentage\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Gender of Victim}}\\
\hspace{1em}Man & 91.35\\
\hspace{1em}Woman & 8.65\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Type of Crime}}\\
\hspace{1em}Murder & 97.82\\
\hspace{1em}Robbery & 2.00\\
\hspace{1em}Body injury followed by death & 0.18\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Ethnicity of Victim}}\\
\hspace{1em}Black and White & 95.11\\
\hspace{1em}Yellow and White & 1.27\\
\hspace{1em}Not Reported & 3.62\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Age of Victim}}\\
\hspace{1em}1 to 12 years & 0.18\\
\hspace{1em}13 to 17 years & 11.31\\
\hspace{1em}18 to 30 years & 61.70\\
\hspace{1em}31 to 65 years & 25.84\\
\hspace{1em}65 years and older & 0.73\\
\hspace{1em}Not Reported & 0.24\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weapon Used}}\\
\hspace{1em}Firearm & 87.51\\
\hspace{1em}Other & 12.49\\
\bottomrule
\end{tabular}
\end{table}

@@

Figure \ref{fig:plot-sed-as-quintiles} displays the geography of
Socio-Economic Deprivation in Recife, after classifying \emph{setores}
by quintiles, whereby the ``High'' class corresponds to \emph{setores}
in the top 20\% of the Socio-Economic Deprivation indicator, and the
``Low'' class corresponds to \emph{setores} in the bottom 20\% of the
Socio-Economic Deprivation indicator. The figure shows a veritable
mosaic of affluence and deprivation, with high deprivation areas
directly in contact, and in some cases almost completely surrounded, by
low deprivation areas. This geographical pattern of unequality, on the
other hand, seems to be characteristic of Brazilian metropolitan
regions, where enclaves of wealth and \emph{favelas} (i.e., urban slums)
can be found in close proximity (see for example Feitosa et al. 2007).

@@

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional
quadrat counts for Lethal and Intentional Violent Crime. The dotted line
indicates the global intensity, simply the total number of criminal
events divided by the area of the region. It can be seen that the
intensity of crime clearly increases with Socio-Economic Deprivation,
and the intensity in areas with the highest levels of Socio-Economic
Deprivation is 208.7065292 \% higher than in places with the lowest
levels of Socio-Economic Deprivation. This is as expected, as research
has consistently found correlations between Socio-Economic Deprivation
and a variety of negative health outcomes, including homicide (Ichihara
et al. 2018). A \(\chi^2\) test of independence on the quadrat counts
yields a value of \(p\leq 0.0001\) (four degrees of freedom), which
comfortably rejects the null hypothesis of homogeneity of intensity of
crime by level of Socio-Economic Deprivation.

@@

Three types of commercial establishments also appear to display
inhomogeneous locational patterns (Table
\ref{tab:quadrat-test-statistics}), with the \(\chi^2\) test of
independence emphaticaly rejecting the null hypothesis. As seen in
Figure \ref{fig:plot-events-quadrat}, ice cream shops tend to be found
more frequently in places with middle-low Socio-Economic Deprivation,
and less frequently in places with middle-high Socio-Economic
Deprivation. Pharmacies show a clearer trend, with locational patterns
that tend to favor places with low Socio-Economic Deprivation. Finally,
supermarkets are found less frequently in places with middle-high and
high Socio-Economic Deprivation.

@@

\caption{\label{tab:table-quadrat-test-statistics}\label{tab:quadrat-test-statistics}$\chi^2$ tests of independence of conditional quadrat counts}
\centering
\begin{tabular}[t]{lccc}
\toprule
Event & Statistic & Degrees of Freedom & p-value\\
\midrule
Catholic & 6.913 & 4 & 0.2811\\
Evangelical & 38.669 & 4 & $<0.001$\\
Spiritist & 9.108 & 4 & 0.1169\\
Ice Crem Shop & 15.108 & 4 & 0.009\\
Pharmacy & 51.869 & 4 & $<0.001$\\
\addlinespace
Supermarket & 23.826 & 4 & $<0.001$\\
\bottomrule
\end{tabular}
\end{table}

@@

The results of the analysis are shown in Figure
\ref{fig:plot-relative-distribution-with-baseline}. The general trends
shown by the relative distribution functions are similar to those seen
above, before the introduction of a baseline function. We notice that at
short distances the intensity of crime as a function of distance to the
various places of worship and commercial establishments is higher than
the baseline intesity (i.e., the intensity according to Socio-Ecomomic
Deprivation). The only exception is intensity as a function of distance
to a Spiritist church, which at short distances is not significantly
different from the baseline intensity (see the confidence bands).

The peak intensity as a function of distance to Evangelical churches is
3.33 times the baseline intensity; for Catholic places of worship, the
highest intesity is 2.12 times the baseline intensity; and the intensity
reaches a peak of 2.67 times the baseline intensity when the covariate
is distance to Spiritist churches. Turning now to inspect the three
types of commercial establishments, we see that the highest intensities
are 2.28, 2.46, and 2.09 times the intensity according to Socio-Economic
Deprivation for ice cream shops, pharmacies, and supermarkets,
respectively.

@@

In this respect, it has been argued that Catholics and some Protestant
groups are more adept at generating bridging social capital with stron
intergroup relations. This can result in greater collective efficacy -
which in turn can decrease the likelihood of crime (see Hoffmann 2015,
2). Along these lines, Beyerlein and Hipp (2005), examined data for
3,157 counties in the United States in 2000, and found that counties
with greater percentages of people affiliated with mainline Protestant
and Catholic traditions tended to have lower crime rates. Similarly,
Warner and Konkel (2019) found that mainline Protestant and briding
churches (but not Catholic churches) correlated positively with informal
social control at the neighborhood level in two cities in Kentucky. Nie
and Yang (2019), in a study of smoking, also found that share of
Catholic population in counties did not reduce the rate of smoking, thus
bringing into question the efficacy of this type of moral community in
generating social control - albeit at a fairly aggregated scale
(counties). Our findings with respect to Catholic places of worship echo
these results at a much higher level of disaggregation, as we fail to
find evidence that Catholic churches project a moral community in space
more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline
Protestants and Catholics, appear to focus less on bridging and more on
bonding social capital with strong within-group relations. Associated
with this, Conservative Protestant doctrine emphasizes individual
responsibility, and therefore social ills tend to be seen as primarily
personal ills in need of religious redemption instead of secular
interventions (Nie and Yang 2019, 2--3). Under this light, a moral
community with such views of social problems may not be effective to
deter crime (Ellison, Burr, and McCall 2003); instead, conservative
Protestant doctrine may be more tolerant of violent behavior when
associated with defense of honor, family, property, or women, or
unfortunate events may be seen as representing legitimate celestial
retribution for moral torpitude. Empirical research by Beyerlein and
Hipp (2005) found that counties with greater percentages of people
affiliated with Evangelical traditions tended to have higher crime
rates. More recent research by Desmond et al.~(2010) in more than 400
block groups in Indianapolis found that neighborhoods with a greater
presence of Evangelical congregations had higher rates of both violent
and property crimes. In the specific context of Brazil, it is important
to consider the role that religion has had in interpreting and
conferring meaning to urban violence, or what Brazilian anthropologist
Patricia Birman has termed ``The Violence of the Just'' (see Birman and
Machado 2012; Birman 2019).

Our analysis suggests that the role of Evangelical churches may be more
complex than thought: on the one hand, our results indicate that Lethal
and Intentional Violent Crime tends to be more intense in the proximity
of Evangelical places of worship, up to a distance of approximately 100
m, but then declines even below the global intensity and the background
intenstiy at distances between 300 m and 100 m. It is at the moment
unclear what could cause this, and candidate explanations could include
the way police respond to crime incidents in different neighborhoods
{[}e.g., Warner, 1997{]}, the likelihood of residents reporting crime to
the police (e.g., Goudriaan, Witterbrood, \& Niewbeerta, 2006), and the
level of opportunity for crimes (e.g., Rountree, Land, \& Miethe, 1994).
Could Evangelicals churches provide greater levels of opportunity for
crime, if adherents to this tradition are seen as meek? And is it
possible that at a different scale, unlike reports by other researchers,
Evangelical churches are more effective at keeping social control? While
these speculations cannot be answered in the context of the present
study, we suggest that these might be fruitful areas for future
research.

Finally, we would like to remark on the results regarding Spiritist
churches, which in some way are the mirror image of those for
Evangelical churches. An extensive scan of the literature finds much
less information about this religious tradition, in particular in the
context of Brazil where it adapted in ways that made it very different
from its original French predecessor (see Graca Arribas 2011). Could it
be that Spiritism is seen by potential criminals as spiritually or
materially risky, if supernatural retribution is feared? Again, we
cannot do more than speculate about this at the moment, but this might
be another worthwhle avenue for future investigation.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Catholic places of worhsip do not seem to geographically project more
  or less of a moral community effect than, say, ice cream shops,
  pharmacies, or supermarkets: the intensity of crime as a function of
  distance to Catholic churches is not significantly different that what
  we observe when we consider distance to any of these types of
  commercial establishments, at any distance.
\item
  The intensity of crime with respect to distance to Evangelical
  churches is more complex than previously known: whereas the intensity
  of crime is higher at relativey short distance from Evangelical places
  of worship, at longer distances it tends to decline even below the
  global intensity of crime and the background level of crime according
  to Socio-Economic Deprivation.
\item
  Spiritist churches are associated with lower intensity of crime at
  very short distance; however, at longer distances(approximately
  between 200 m and 500 m), the intensity of crime is in fact higher
  than the intensity of crime at comparable distances from Catholic and
  Evangelical places of worship, or any of the three kinds of commercial
  establishments.
\end{enumerate}

@@

In summary, the research presented on this paper provides information
about the potential of different types of moral communities to reduce
crime, and should be of interest to policy makers as they assess whether
formal or informal forms of social control can be effective to deter
criminal behavior.

@@

@@

@@

@@

@@

@@

@@

>>>>

This concern is not idle: the endemic malady of violent crime represents
an important factor that threatens to derail sustainable development
goals. Violent crime, furthermore, tends to be more acute in those
regions most in need of development. On the one hand, violent crime
represents an economic and social drag (Becker and Kassouf 2017). The
impacts are multifaceted, since they affect people's well-being, either
through the loss of human life, mental health, and limitations in the
right to public spaces (Doran and Burgess 2011), or through disturbances
in schooling and academic achievement (United Nations Office on Drugs
and Crime 2019b). In turn, these negative effects combine to create the
unfortunate conditions that tend to breed criminal behavior, thus
creating a vicious cycle of economic disadvantage and violence (United
Nations Office on Drugs and Crime 2019b). It is not surprising, given
these peculiarities, that researchers have joined calls for increased
attention to the study of the patterns of violent crime in Low and
Middle Income Countries (LMIC; Murray, Cerqueira, and Kahn 2013).

Added to the scenario, formal social control institutions in countries
like Brazil leave something (or much) to be desired: while in practice
they should inhibit criminal behavior, they suffer from deep
deficiencies that end up reducing the deterrent power of the justice
system. Serious institutional problems include the inefficiency of the
police, the lack of national legislation, the glacial pace of judicial
processes, and the weak situation of the prison system in the country
(Menezes et al. 2013). This makes it even more urgent to understand the
role of \emph{informal social controls}, i.e., the ability of community
organizations such as schools, clubs, and neighborhood associations to
suppress crime by strengthening the capabilities of neighbors to control
inappropriate behavior (Groff 2015).

In order to understand the factors that can potentially help to deter
crime, beyond formal control, it is important to identify empirical
regularities. Criminological factors include concrete elements (such as
the presence of arms or drugs, or elements of the built environment),
and also figurative factors, which include the social cost of deviant
behavior. Accordingly, a number of studies have investigated various
aspects of the environment and neighborhood design (e.g., Foster,
Giles-Corti, and Knuiman 2010; He, Páez, and Liu 2017; Loukaitou-Sideris
et al. 2001), whereas other studies have focused on exposure to
environmental attributes that signal weakened norms, such as liquor and
tobacco outlets (e.g., Brower and Carroll 2007; Deryol et al. 2016;
Lipton et al. 2008; Quick, Law, and Luan 2017).

@@

With the above considerations in mind, the objective of this paper is to
investigate whether and how places of worship correlate spatially with
criminal events, or in other words, to investigate whether their signals
as moral communities have a discernible spatial dimension. The case
study is the city of Recife in the state of Pernambuco, in Brazil's
Northeast. Recife is a large and important metropolitan area in a
historically poor region, and afflicted by high levels of violent crime,
having the dubious honor of being one of five state capitals with the
highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

@@

An early effort to theorize the effect of religion on criminal behavior
was the \emph{hellfire hypothesis} of Hirschi and Stark (1969). The
focus of this hypothesis is the threat of extra-temporal (and possibly
eternal) punishment, and how this threat can deter believers from
commandment-breaking (also see Pascal's wager). Accordingly, Hirsch and
Stark (1969) posit that negative correlations between crime and religion
are a consequence of a sense of commitment with normative values, a
commitment that is ritually reinforced in a regular fashion (e.g., the
liturgical rite of peace in Catholic Mass). Rohrbaugh and Jessor (1975),
for instance, note that by attributing to divinity (or some other
supra-terrenal entity) the supreme force of punishment, religion helps
to build a sense of obduracy, or ``hardening'' against temptation.

@@

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plausibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization theory are
thought to occur at relatively small geographical scales.

@@

Using quadrats of equal size is a convenient simplification, but in
principle the areas could be different - in which case the count of
events would be proportional to the area of the quadrat under
homogeneity. An interesting variation of this technique is conditional
quadrat counts, whereby the partition of regions
\(\mathcal{A}_1,\cdots,\mathcal{A}_m\) is done to reflect an underlying
covariate of interest, say \(Z\). By introducing a covariate as a
partitioning criterion, it becomes possible to calculate estimators of
the intensity of quadrats at different levels of the value of the
covariate. For exploratory purposes, we can plot average intensity by
quadrat, and compare to the global intensity.

@@

\caption{\label{tab:table-descriptive-statistics}\label{tab:descriptive-statistics}Characteristics of Violent Crime in Recife, July, 2008 - June, 2009}
\centering
\begin{tabular}[t]{lr}
\toprule
 & Percentage\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Gender of Victim}}\\
\hspace{1em}Man & 91.35\\
\hspace{1em}Woman & 8.65\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Type of Crime}}\\
\hspace{1em}Murder & 97.82\\
\hspace{1em}Robbery & 2.00\\
\hspace{1em}Body injury followed by death & 0.18\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Ethnicity of Victim}}\\
\hspace{1em}Black and Brown & 95.11\\
\hspace{1em}Yellow and White & 1.27\\
\hspace{1em}Not Reported & 3.62\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Age of Victim}}\\
\hspace{1em}1 to 12 years & 0.18\\
\hspace{1em}13 to 17 years & 11.31\\
\hspace{1em}18 to 30 years & 61.70\\
\hspace{1em}31 to 65 years & 25.84\\
\hspace{1em}65 years and older & 0.73\\
\hspace{1em}Not Reported & 0.24\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weapon Used}}\\
\hspace{1em}Firearm & 87.51\\
\hspace{1em}Other & 12.49\\
\bottomrule
\end{tabular}
\end{table}

@@

Figure \ref{fig:plot-sed-as-quintiles} displays the geography of
Socio-Economic Deprivation in Recife, after classifying \emph{setores}
by quintiles, whereby the ``High'' class corresponds to \emph{setores}
in the top 20\% of the Socio-Economic Deprivation indicator, and the
``Low'' class corresponds to \emph{setores} in the bottom 20\% of the
Socio-Economic Deprivation indicator. The figure shows a veritable
mosaic of affluence and deprivation, with high deprivation areas
directly in contact, and in some cases almost completely surrounded, by
low deprivation areas. This geographical pattern of inequality, on the
other hand, seems to be characteristic of Brazilian metropolitan
regions, where enclaves of wealth and \emph{favelas} (i.e., urban slums)
can be found in close proximity (see for example Feitosa et al. 2007).

@@

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional
quadrat counts for Lethal and Intentional Violent Crime. The dotted line
indicates the global intensity, simply the total number of criminal
events divided by the area of the region. It can be seen that the
intensity of crime clearly increases with Socio-Economic Deprivation,
and the intensity in areas with the highest levels of Socio-Economic
Deprivation is 208.71\% higher than in places with the lowest levels of
Socio-Economic Deprivation. This is as expected, as research has
consistently found correlations between Socio-Economic Deprivation and a
variety of negative health outcomes, including homicide (Ichihara et al.
2018). A \(\chi^2\) test of independence on the quadrat counts yields a
value of \(p\leq 0.0001\) (four degrees of freedom), which comfortably
rejects the null hypothesis of homogeneity of intensity of crime by
level of Socio-Economic Deprivation.

@@

Three types of commercial establishments also appear to display
inhomogeneous locational patterns (Table
\ref{tab:quadrat-test-statistics}), with the \(\chi^2\) test of
independence emphatically rejecting the null hypothesis. As seen in
Figure \ref{fig:plot-events-quadrat}, ice cream shops tend to be found
more frequently in places with middle-low Socio-Economic Deprivation,
and less frequently in places with middle-high Socio-Economic
Deprivation. Pharmacies show a clearer trend, with locational patterns
that tend to favor places with low Socio-Economic Deprivation. Finally,
supermarkets are found less frequently in places with middle-high and
high Socio-Economic Deprivation.

@@

\caption{\label{tab:table-quadrat-test-statistics}\label{tab:quadrat-test-statistics}$\chi^2$ tests of independence of conditional quadrat counts}
\centering
\begin{tabular}[t]{lccc}
\toprule
Event & Statistic & Degrees of Freedom & p-value\\
\midrule
Catholic & 6.913 & 4 & 0.2811\\
Evangelical & 38.669 & 4 & $<0.001$\\
Spiritist & 9.108 & 4 & 0.1169\\
Ice Cream Shop & 15.108 & 4 & 0.009\\
Pharmacy & 51.869 & 4 & $<0.001$\\
\addlinespace
Supermarket & 23.826 & 4 & $<0.001$\\
\bottomrule
\end{tabular}
\end{table}

@@

The results of the analysis are shown in Figure
\ref{fig:plot-relative-distribution-with-baseline}. The general trends
shown by the relative distribution functions are similar to those seen
above, before the introduction of a baseline function. We notice that at
short distances the intensity of crime as a function of distance to the
various places of worship and commercial establishments is higher than
the baseline intensity (i.e., the intensity according to Socio-Economic
Deprivation). The only exception is intensity as a function of distance
to a Spiritist church, which at short distances is not significantly
different from the baseline intensity (see the confidence bands).

The peak intensity as a function of distance to Evangelical churches is
3.33 times the baseline intensity; for Catholic places of worship, the
highest intensity is 2.12 times the baseline intensity; and the
intensity reaches a peak of 2.67 times the baseline intensity when the
covariate is distance to Spiritist churches. Turning now to inspect the
three types of commercial establishments, we see that the highest
intensities are 2.28, 2.46, and 2.09 times the intensity according to
Socio-Economic Deprivation for ice cream shops, pharmacies, and
supermarkets, respectively.

@@

In this respect, it has been argued that Catholics and some Protestant
groups are more adept at generating bridging social capital with strong
intergroup relations. This can result in greater collective efficacy -
which in turn can decrease the likelihood of crime (see Hoffmann 2015,
2). Along these lines, Beyerlein and Hipp (2005), examined data for
3,157 counties in the United States in 2000, and found that counties
with greater percentages of people affiliated with mainline Protestant
and Catholic traditions tended to have lower crime rates. Similarly,
Warner and Konkel (2019) found that mainline Protestant and bridging
churches (but not Catholic churches) correlated positively with informal
social control at the neighborhood level in two cities in Kentucky. Nie
and Yang (2019), in a study of smoking, also found that share of
Catholic population in counties did not reduce the rate of smoking, thus
bringing into question the efficacy of this type of moral community in
generating social control - albeit at a fairly aggregated scale
(counties). Our findings with respect to Catholic places of worship echo
these results at a much higher level of disaggregation, as we fail to
find evidence that Catholic churches project a moral community in space
more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline
Protestants and Catholics, appear to focus less on bridging and more on
bonding social capital with strong within-group relations. Associated
with this, Conservative Protestant doctrine emphasizes individual
responsibility, and therefore social ills tend to be seen as primarily
personal ills in need of religious redemption instead of secular
interventions (Nie and Yang 2019, 2--3). Under this light, a moral
community with such views of social problems may not be effective to
deter crime (Ellison, Burr, and McCall 2003); instead, conservative
Protestant doctrine may be more tolerant of violent behavior when
associated with defense of honor, family, property, or women, or
unfortunate events may be seen as representing legitimate celestial
retribution for moral turpitude. Empirical research by Beyerlein and
Hipp (2005) found that counties with greater percentages of people
affiliated with Evangelical traditions tended to have higher crime
rates. More recent research by Desmond et al.~(2010) in more than 400
block groups in Indianapolis found that neighborhoods with a greater
presence of Evangelical congregations had higher rates of both violent
and property crimes. In the specific context of Brazil, it is important
to consider the role that religion has had in interpreting and
conferring meaning to urban violence, or what Brazilian anthropologist
Patricia Birman has termed ``The Violence of the Just'' (see Birman and
Machado 2012; Birman 2019).

Our analysis suggests that the role of Evangelical churches may be more
complex than thought: on the one hand, our results indicate that Lethal
and Intentional Violent Crime tends to be more intense in the proximity
of Evangelical places of worship, up to a distance of approximately 100
m, but then declines even below the global intensity and the background
intensity at distances between 300 m and 100 m. It is at the moment
unclear what could cause this, and candidate explanations could include
the way police respond to crime incidents in different neighborhoods,
the likelihood of residents reporting crime to the police, and the level
of opportunity for crimes (Warner and Konkel 2019). Could Evangelicals
churches provide greater levels of opportunity for crime, if adherents
to this tradition are seen as meek? And is it possible that at a
different scale, unlike reports by other researchers, Evangelical
churches are more effective at keeping social control? While these
speculations cannot be answered in the context of the present study, we
suggest that these might be fruitful areas for future research.

Finally, we would like to remark on the results regarding Spiritist
churches, which in some way are the mirror image of those for
Evangelical churches. An extensive scan of the literature finds much
less information about this religious tradition, in particular in the
context of Brazil where it adapted in ways that made it very different
from its original French predecessor (see Graca Arribas 2011). Could it
be that Spiritism is seen by potential criminals as spiritually or
materially risky, if supernatural retribution is feared? Again, we
cannot do more than speculate about this at the moment, but this might
be another worthwhile avenue for future investigation.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Catholic places of worship do not seem to geographically project more
  or less of a moral community effect than, say, ice cream shops,
  pharmacies, or supermarkets: the intensity of crime as a function of
  distance to Catholic churches is not significantly different that what
  we observe when we consider distance to any of these types of
  commercial establishments, at any distance.
\item
  The intensity of crime with respect to distance to Evangelical
  churches is more complex than previously known: whereas the intensity
  of crime is higher at relatively short distance from Evangelical
  places of worship, at longer distances it tends to decline even below
  the global intensity of crime and the background level of crime
  according to Socio-Economic Deprivation.
\item
  Spiritist churches are associated with lower intensity of crime at
  very short distance; however, at longer distances(approximately
  between 200 m and 500 m), the intensity of crime is in fact higher
  than the intensity of crime at comparable distances from Catholic and
  Evangelical places of worship, or any of the three kinds of commercial
  establishments.
\end{enumerate}

@@

In summary, the research presented on this paper provides information
about the potential of different types of moral communities to reduce
crime, and should be of interest to policy makers as they assess whether
formal or informal forms of social control can be effective to deter
criminal behavior, in order to achieve development goals.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

The following \texttt{R} packages were used in the course of this
investigation and the authors wish to acknowledge their developers:
\texttt{faterize} (Ross 2018), \texttt{ggthemes} (Arnold 2018),
\texttt{kableExtra} (Zhu 2018), \texttt{knitr} (Xie 2015, 2018),
\texttt{maptools} (Bivand and Lewin-Koh 2019), \texttt{spatstat}
(Baddeley and Turner 2005), \texttt{raster} (Hijmans 2018),
\texttt{rticles} (Allaire et al. 2018), \texttt{sf} (Pebesma 2018), and
\texttt{tidyverse} (Wickham 2017).

@@

\leavevmode\hypertarget{ref-Allaire2018rticles}{}%
Allaire, JJ, Yihui Xie, R Foundation, Hadley Wickham, Journal of
Statistical Software, Ramnath Vaidyanathan, Association for Computing
Machinery, et al. 2018. \emph{Rticles: Article Formats for R Markdown}.
\url{https://CRAN.R-project.org/package=rticles}.

\leavevmode\hypertarget{ref-Arnold2018}{}%
Arnold, Jeffrey B. 2018. \emph{Ggthemes: Extra Themes, Scales and Geoms
for 'Ggplot2'}. \url{https://CRAN.R-project.org/package=ggthemes}.

@@

\leavevmode\hypertarget{ref-Baddeley2005spatstat}{}%
Baddeley, Adrian, and Rolf Turner. 2005. ``spatstat: An R Package for
Analyzing Spatial Point Patterns.'' \emph{Journal of Statistical
Software} 12 (6): 1--42. \url{http://www.jstatsoft.org/v12/i06/}.

@@

\leavevmode\hypertarget{ref-Bivand2019maptools}{}%
Bivand, Roger, and Nicholas Lewin-Koh. 2019. \emph{Maptools: Tools for
Handling Spatial Objects}.
\url{https://CRAN.R-project.org/package=maptools}.

@@

\leavevmode\hypertarget{ref-Hijmans2018raster}{}%
Hijmans, Robert J. 2018. \emph{Raster: Geographic Data Analysis and
Modeling}. \url{https://CRAN.R-project.org/package=raster}.

@@

\leavevmode\hypertarget{ref-Pebesma2018simple}{}%
Pebesma, Edzer. 2018. ``Simple Features for R: Standardized Support for
Spatial Vector Data.'' \emph{The R Journal} 10 (1): 439--46.
\url{https://doi.org/10.32614/RJ-2018-009}.

@@

\leavevmode\hypertarget{ref-Ross2018fasterize}{}%
Ross, Noam. 2018. \emph{Fasterize: Fast Polygon to Raster Conversion}.
\url{https://CRAN.R-project.org/package=fasterize}.

@@

\leavevmode\hypertarget{ref-Wickham2017}{}%
Wickham, Hadley. 2017. \emph{Tidyverse: Easily Install and Load the
'Tidyverse'}. \url{https://CRAN.R-project.org/package=tidyverse}.

\leavevmode\hypertarget{ref-Xie2015}{}%
Xie, Yihui. 2015. \emph{Dynamic Documents with R and Knitr}. 2nd ed.
Boca Raton, Florida: Chapman; Hall/CRC. \url{https://yihui.name/knitr/}.

\leavevmode\hypertarget{ref-Xie2018}{}%
---------. 2018. \emph{Knitr: A General-Purpose Package for Dynamic
Report Generation in R}. \url{https://yihui.name/knitr/}.

\leavevmode\hypertarget{ref-Zhu2018}{}%
Zhu, Hao. 2018. \emph{KableExtra: Construct Complex Table with 'Kable'
and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.
